Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods

### 2022-09-26 inst/surfing_jsa_rebuttal_minorcomments.tex

Page 4: The authors mention the surfer Vries---who does not appear to
be a wildcard---whose maximum likelihood strength is zero. Why is
this? My guess is that Vries has never won a competitive surfing heat
(and therefore there is no information on his true, underlying
ability), however a brief clarification here woud be useful.

\textcolor{blue}{The referee is quite right: Vries competed only twice
  in 2019, coming last (second and third) in each of his two sessions.
  Brief explanation added to he mauscript.}\\

Page 6 / Figure 1: I'm pleased to see the authors replace the old pie
chart in Figure 1 with a dot chart---this is a much clearer
visualization of the results.  On Page 6 the authors state: ``Figure 1
shows the maximum likelihood estimate for the 23 competitors'
strengths, which appears to show a wide range from Florence at about
$17.5\%$ down to Morais at about $1\%$". What exactly do the
quantities $17.5\%$ and $1\%$ refer to/how are they computed? A quick
clarification of these quantities would help explain the practical
implications of the findings.

\textcolor{blue}{The figures are, as stated, the maximum likelihood
  estimates of the competitors' strengths.  We think that our use of
  percentages (which add to 100) instead of strengths (which add to 1)
  was the cause of the confusion and we have amended the manuscript to
  use strengths, not percentages.}\\

@@

Page 7/Figure 2: The authors suggest there is a loose positive
correlation between maximum likelihood strength and total points,
which is undoubtedly true. However, the x-axis on Figure 2 itself has
been distorted, which makes the trend appear more linear than it
really is. I'd prefer to see the x-axis undistorted, unless the
authors are explicitly trying to make a point about the trend being
log-linear, in which case I'd like to see this briefly discussed.

\textcolor{blue}{Horizontal axis now linear; no other change except
  for tiny movement of labels so they do not obscure one another.  We
  are mortified to see that the linearity is more apparent without the
  log-transform.}\\

@@

Page 7: The authors suggest we may reject the hypothesis that the
strength of the entity exceeds about 0.025. Does this assertation come
with a supporting p-value? It is unclear from Figure 4 how this
conclusion is reached.

\textcolor{blue}{This was a (devil-inspired) typo.  The value should
  read 0.015.  The reasoning was that at 0.015 the support was -2.  We
  have ddded a sentence of explanation on p7 and also the figure
  caption.}\\

@@

Page 17: Figure 2 caption is a little brief and could better explain
what is observed in the plot (e.g. the positive correlation between
variables). Especially as the authors have distorted the x-axis and
fitted a linear trend through the data. What is the reason for this?
As mentioned in a previous comment, I’d prefer to see the x-axis in
its raw form, unless there is a reason it has been distorted.

\textcolor{blue}{More detail added in caption}\\

Page 18: Figure 3 title – this doesn’t provide the reader with any
additional information. It does seem a little odd that Figure 2 does
not have a title, Figure 3 does, and Figure 4 does not. For
consistency, the authors could simply remove the title from Figure 33,
as it does not really add any value.

\textcolor{blue}{Done (title removed from figure 3 as suggested)}\\

@@

Page 19: Figure 4 Capitalisation of y-axis title.  General comment:
should the Table and Figure captions have a full-stop at the end of
the last sentence?

\textcolor{blue}{Done.  Looking at recent JSA articles the captions do
  not have full-stops and we just copied that style}\\

@@

General comment: As part of a potential future publication in the
field, I'd be curious to see the methodology in Section 7 applied to
World Surf League data from events that occurred prior to the
announcement that surfing would be in the Olympics.  Comparing these
results with those in the paper that use data from after this
announcement (perhaps supplemented with additional data and results
from post-2019), would help in our understanding whether there truly
has been an observable change in non-competitive behaviour between
compatriots.  Any such analysis is clearly well beyond the realm of
minor revisions and is not something I expect of the authors prior to
final submission; however, I thought it could be an interesting next
step for the application of the methodology outlined in the paper!

\textcolor{blue}{This suggestion is an excellent one (which had not
  occurred to us); as remarked, it is not in scope for the current
  paper.  We do plan to work on this idea in the future and have added
  a brief discussion to the ``Future work'' section/.
}

### 2022-09-26 inst/surfing_jsa.tex
very minor typesetting pursuant to issue #166

Table~\ref{rawresults} shows the first few and last few entries in the
dataset we use, drawn from 2019 WSL.  The first line shows that, at
one of the heats in WCT01 (Coolangatta, QLD), Colapinto came first,
Bailey second, and Wright came third: we write
$\mbox{Colapinto}\succ\mbox{Bailey}\succ\mbox{Wright}$.  The last line
shows that, at WCT11 (Banzai Pipeline), Ferreira came first and Medina
second: $\mbox{Ferreira}\succ\mbox{Medina}$.  We can convert this
dataset into a Plackett-Luce likelihood function but first have to
remove the wildcards whose strength is typically very low.  We also
remove De Vries who, while not a wildcard, competed only twice in 2019
and came last each time, thus having zero maximum likelihood strength.
For the first line we would have Plackett-Luce likelihood function as

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\citep{hankin2010}; package idiom
is to use {\tt equalp.test()} which reports its findings in a
standardised format.  This gives us a support difference of about
0.05, well short of Edwards's 2 units of support criterion.
Alternatively, we could observe that the
$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
assert that there is no strong evidence for Ferreira's actually having
a competitive advantage over Medina.  We may similarly wonder whether
Medina is equal in strength to Wilson; here we reject the null with
support difference of about 2.33, corresponding to a $p$-value of
$0.031$.

Medina's performance was startlingly good at Freshwater Pro ({\tt
  wct08}).  Indeed, Medina came first at only two venues (the other
was at Jeffries Bay, South Africa) and it is natural to enquire
whether there is statistical evidence to support the assertion that he
performed better under the more controlled conditions of the Lemoore
Surf Ranch.  To do this, we create an artificial surfer, {\tt
  Medina\_wct08}, who competes at Lemoore, and test the null that {\tt
  Medina} has the same competitive strength as {\tt Medina\_wct08}.
We reject this null at a $p$-value of 0.043 and conclude that Medina's
performance at Lemoore is significantly better than his performance at
natural venues.  From an inferential perspective, this test is
interesting because competitor {\tt Medina} never faces {\tt
  Medina\_wct08}.  It is as though competitor A consistently beats B,
and B consistently beats C in pairwise competition, but A has never
directly competed against C.  Nevertheless we may assert with some
confidence that the Bradley-Terry strength of A exceeds that of C in
the absence of nontransitive effects~\citep{west2008}, via indirect
comparison.

@@

We analysed results from the 2019 WSL using reified
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Reified Bradley-Terry allows the use of
formal statistical methods in ways that analysing points scored does
not.  For example, we can be confident that the results reflect
differences in skill (rather than random variation); there is no
statistical difference in strength between the top two officially
ranked competitors, Medina and Ferreira; but we may reject the
hypothesis that Medina and Wilson are equal in strength.

We discuss and assess the ``Brazilian storm'' phenomenon using reified
Bradley-Terry techniques, and show that there is no strong evidence to
support the supposed Brazilian preference for point and beach break
beaches.  We also suggest that surfing's being included in the 2020
Summer Olympics might result in modified strategies to bolster
intranational ranking at the expense of WCT performance; we quantify
the strength of such a phenomenon as the strength of a reified
Bradley-Terry entity: a ``compatriot monster'' with an estimated
strength of about $0.017$.

@@

% use   \begin{tabular}{@{}c@{}}\small{WCT} \\ 01\end{tabular} for WCT\\01
\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality & \multirow{ 5}{*}{1} &  &  & WCT & & &points\\
            &             &  01   & 02& 03& \ldots& 10 & 11 & \\ \hline
 Ferreira   &   BRA       &   1   &   5   &   17  & \ldots  &   1   &   1   &   59740\\
 Medina     &   BRA       &   5   &   5   &   17  & \ldots  &   9   &   2   &   56475\\
 Smith 	    &   ZAF       &   3   &   3   &   17  & \ldots  &   2   &   17  &   49985\\
 Toledo     &   BRA       &   9   &   2   &   5   & \ldots  &   5   &   17  &   49145\\
 Andino     &   USA       &   2   &   17  &   5   & \dots   &   5   &   9   &   46655\\
 $\vdots$   &             &   &   &       &       & $\vdots$&       &       &
\end{tabular}
\caption{Conventional league table for the top five competitors in WSL
  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
  Ferreira coming first, Andino second and Smith third}
\end{table}

>>>>

Table~\ref{rawresults} shows the first few and last few entries in the
dataset we use, drawn from 2019 WCT.  The first line shows that, at
one of the heats in WCT01 (Coolangatta, QLD), Colapinto came first,
Bailey second, and Wright came third: we write
$\mbox{Colapinto}\succ\mbox{Bailey}\succ\mbox{Wright}$.  The last line
shows that, at WCT11 (Banzai Pipeline), Ferreira came first and Medina
second: $\mbox{Ferreira}\succ\mbox{Medina}$.  We can convert this
dataset into a Plackett-Luce likelihood function but first have to
remove the wildcards whose strength is typically very low.  We also
remove De Vries who, while not a wildcard, competed only twice in 2019
and came last each time, thus having zero maximum likelihood strength.
For the first line we would have a Plackett-Luce likelihood function of

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\citep{hankin2010}; package idiom
is to use {\tt equalp.test()} which reports its findings in a
standardised format.  This gives a support difference of about
0.05, well short of Edwards's 2 units of support criterion.
Alternatively, we could observe that~$\mathcal{S}=
-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
assert that there is no strong evidence for Ferreira actually having
a competitive advantage over Medina.  We may similarly wonder whether
Medina is equal in strength to Wilson; here we reject the null with
support difference of about 2.33, corresponding to a $p$-value of
$0.031$.

Medina's performance was startlingly good at Freshwater Pro (WCT08).
Indeed, Medina came first at only two venues (the other
was at Jeffries Bay, South Africa) and it is natural to enquire
whether there is statistical evidence to support the assertion that he
performed better under the more controlled conditions of the Lemoore
Surf Ranch.  To do this, we create an artificial surfer, {\tt
  Medina\_wct08}, who competes at Lemoore, and test the null that {\tt
  Medina} has the same competitive strength as {\tt Medina\_wct08}.
We reject this null at a $p$-value of 0.043 and conclude that Medina's
performance at Lemoore is significantly better than his performance at
natural venues.  From an inferential perspective, this test is
interesting because competitor {\tt Medina} never faces {\tt
  Medina\_wct08}.  It is as though competitor A consistently beats B,
and B consistently beats C in pairwise competition, but A has never
directly competed against C.  Nevertheless we may assert with some
confidence that the Bradley-Terry strength of A exceeds that of C in
the absence of nontransitive effects~\citep{west2008}, via indirect
comparison.

@@

We analysed results from the 2019 WCT using reified
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Reified Bradley-Terry allows the use of
formal statistical methods in ways that analysing points scored does
not.  For example, we can be confident that the results reflect
differences in skill (rather than random variation); there is no
statistical difference in strength between the top two officially
ranked competitors, Medina and Ferreira; but we may reject the
hypothesis that Medina and Wilson are equal in strength.

We discuss and assess the ``Brazilian storm'' phenomenon using reified
Bradley-Terry techniques, and show that there is no strong evidence to
support the supposed Brazilian preference for point and beach break
wave types.  We also suggest that surfing's being included in the 2020
Summer Olympics might result in modified strategies to bolster
intranational ranking at the expense of WCT performance; we quantify
this phenomenon by introducing a reified
Bradley-Terry entity: a ``compatriot monster'' with an estimated
strength of about $0.017$.

@@

% use   \begin{tabular}{@{}c@{}}\small{WCT} \\ 01\end{tabular} for WCT\\01
\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality & \multirow{ 5}{*}{1} &  &  & WCT & & &points\\
            &             &  01   & 02& 03& \ldots& 10 & 11 & \\ \hline
 Ferreira   &   BRA       &   1   &   5   &   17  & \ldots  &   1   &   1   &   59740\\
 Medina     &   BRA       &   5   &   5   &   17  & \ldots  &   9   &   2   &   56475\\
 Smith 	    &   ZAF       &   3   &   3   &   17  & \ldots  &   2   &   17  &   49985\\
 Toledo     &   BRA       &   9   &   2   &   5   & \ldots  &   5   &   17  &   49145\\
 Andino     &   USA       &   2   &   17  &   5   & \dots   &   5   &   9   &   46655\\
 $\vdots$   &             &   &   &       &       & $\vdots$&       &       &
\end{tabular}
\caption{Conventional league table for the top five competitors in the WCT
  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
  Ferreira coming first, Andino second and Smith third}
\end{table}
